---
title: CV Analyses
author: Thomas Ward
date: "`r Sys.Date()`"
output: pdf_document
---


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(irr)
library(showtext)
font_add_google("Lato")
showtext_auto()
theme_set(theme_classic(base_family = "Lato"))
```

# Load results

```{r}
results <- read_csv("../data/cv_results.csv", col_types = "ciiiii")
```

# Utility functions

## Calculate Krippendorff's alpha

Convenience function that can be used with `summarise()`

```{r}
kripp_alpha <- function(col1, col2) {
    matrix(c(col1, col2), ncol = length(col1), byrow = TRUE) %>%
        irr::kripp.alpha(method = "ordinal") %>%
        pluck("value")
}
```

## Calculate Krippendorff's alpha in boostrap

This will be used by dplyr in nested list dataframes:

```{r}
kripp_alpha_df <- function(df) {
    df %>%
        as.matrix() %>%
        t() %>%
        irr::kripp.alpha(method = "ordinal") %>%
        pluck("value")
}
```

# Computer Vision Models Cross-Validated Performance

Below I will calculate the cross-validated performance of the two CV models.

The first is `pgs_combo`,
which is the PGS determined by taking the results of the adhesion and appearance networks and
calculating the PGS.

The second is `pgs_only`, which is a network that was trained to only
classify PGS, not taking into account any of the subcomponents.

```{r}
per_fold_results  <- results %>%
    select(name, gt, pgs_combo, pgs_only, fold) %>%
    group_by(fold) %>%
    summarise(across(c("pgs_combo", "pgs_only"), list(krippa = ~ kripp_alpha(., gt))))
knitr::kable(per_fold_results, digits = 3, caption = "Per-fold metrics")
```

Now that we have per-fold metrics, we can calculate CV performance metrics,
including mean, standard deviation (sd), and standard error (se):

```{r}
per_fold_results %>%
    select(-fold) %>%
    summarise(
        across(everything(),
            list(mean = mean, sd = sd, se = ~ sd(.) / sqrt(n()))
        )
    ) %>%
    # make into nice table
    pivot_longer(
        everything(),
        names_to = c("user", NA, "statistic", ".value"),
        names_pattern = "pgs_([a-z]+)_(sqerr_)?([a-z]+)_([a-z]+)"
    ) %>%
    # calc 95% intervals using central limit theorem
    rowwise() %>%
    mutate(
        conf.low = mean - 1.96 * se,
        conf.high = mean + 1.96 * se
    ) %>%
    ungroup() %>%
    arrange(statistic, user) %>%
    knitr::kable(
        digits = 2,
        caption = "Cross-Validated Krippendorff's alpha performance"
    )
```

# Second surgeon performance

The second surgeon annotated all representative frames in one go.
As recommended by Krippendorff,
we can calculate the alpha statistic,
then use the bootstrap to calculate confidence intervals.


```{r}
set.seed(1234)
surg2_metrics <- results %>%
    select(gt, pgs_surg2) %>%
    modelr::bootstrap(10000) %>%
    transmute(
        kripa = map_dbl(strap, compose(kripp_alpha_df, as.data.frame))
    ) %>%
    summarise_all(
        list(
            mean = mean,
            low_ci = ~ quantile(., probs = c(0.025)),
            high_ci = ~ quantile(., probs = c(0.975))
        )
    ) %>%
    pivot_longer(everything(), names_to = "statistic")
```

```{r}
surg2_metrics %>%
    knitr::kable(
        digits = 2,
        caption = "Performance of 2nd Surgeon, CI calculated with the bootstrap"
    )
```

# Confusion Matrices

```{r}
calculate_confusion_matrix <- function(df, groundtruth, prediction) {
    df %>%
        select(gt = {{ groundtruth }}, pred = {{prediction}}) %>%
        count(gt, pred) %>%
        group_by(gt) %>%
        mutate(prop = n / sum(n)) %>%
        ungroup() %>%
        mutate(gt = fct_rev(as.factor(gt)))
}
```

```{r}
# df and specify groundtruth col and prediction col
plot_confusion_matrix <- function(df, title = "Confusion Matrix") {
    df %>%
        ggplot(aes(pred, gt)) +
        scale_fill_gradient(name = "Proportion\n", low = "white", high = "#000099", limits = c(0, 1)) +
        geom_tile(aes(fill = prop)) +
        geom_text(aes(label = n)) +
        theme_classic() +
        theme(
            axis.ticks = element_blank(),
            axis.line = element_blank()
        ) +
        labs(
            x = "PGS Classification",
            y = "Ground Truth PGS",
            title = title
        )
}
```

```{r}
only_results <- calculate_confusion_matrix(results, gt, pgs_only)
combo_results <- calculate_confusion_matrix(results, gt, pgs_combo)
surg2_results <- calculate_confusion_matrix(results, gt, pgs_surg2)
```

```{r}
only_cf <- plot_confusion_matrix(only_results, "PGS only CNN") +
    theme(legend.position = "none")
combo_cf <- plot_confusion_matrix(
        combo_results,
        "PGS from adhesion/\nappearance CNN"
    ) +
    theme(legend.position = "none")
surg2_cf <- plot_confusion_matrix(surg2_results, "Second Surgeon PGS") +
    theme(legend.position = "none")
```

```{r}
library(patchwork)
(only_cf + plot_spacer() + combo_cf ) / (plot_spacer() + surg2_cf + plot_spacer())
```

## Paper figure

```{r}
bind_rows(
        mutate(only_results, network = "PGS-only"),
        mutate(combo_results, network = "PGS-combo")
    ) %>%
    mutate(network = parse_factor(network, levels = c("PGS-only", "PGS-combo"))) %>%
    plot_confusion_matrix(title = "") +
    facet_wrap(vars(network), ncol = 2)
```

```{r}
ggsave("../output/confusion_matrices.pdf", width = 10, height = 6)
ggsave("../output/confusion_matrices.svg", width = 10, height = 6)
```

# Environment

```{r}
sessioninfo::session_info()
```
