---
title: Prep Folds
author: Thomas Ward
date: "`r Sys.Date()`"
output: pdf_document
---

This document will prep the cross-validation folds from the ground truth data
into CSVs appropriate for fastai's csv dataloader.

```{r}
library(tidyverse)
```

# Prepare data

## Generate splits stratified by PGS

```{r}
pgs_gt <- read_csv("../data/pgs_gt.csv", col_types = "cc")
```

```{r}
nsplits = 10
pgs_splits <- pgs_gt %>%
    group_by(pgs) %>%
    mutate(data_split = ntile(n = nsplits)) %>%
    ungroup()
```

## Load appearance and adhesions ground truth

```{r}
appearance_gt <- read_csv("../data/appearance_gt.csv", col_types = "cc")
adhesions_gt <- read_csv("../data/adhesions_gt.csv", col_types = "cc")
```

## Add PGS stratified splits info to appearance and adhesions

```{r}
appearance_splits <- appearance_gt %>%
    inner_join(pgs_splits, by = "fname") %>%
    select(-pgs)
```

```{r}
adhesions_splits <- adhesions_gt %>%
    inner_join(pgs_splits, by = "fname") %>%
    select(-pgs)
```

# Write data for fastai

The fastai csv reader data loader wants three columns:

1. filename
2. label
3. Boolean indicator for validation set or not (fastai is in Python, so "True" and "False")

```{r}
for (i in 1:nsplits) {
    appearance_splits %>%
        mutate(is_valid = if_else(data_split == i, "True", "False"), .keep = "unused") %>%
        write_csv(paste0("../output/appearances_split_", i, ".csv"))
    adhesions_splits %>%
        mutate(is_valid = if_else(data_split == i, "True", "False"), .keep = "unused") %>%
        write_csv(paste0("../output/adhesions_split_", i, ".csv"))
    pgs_splits %>%
        mutate(is_valid = if_else(data_split == i, "True", "False"), .keep = "unused") %>%
        write_csv(paste0("../output/pgs_split_", i, ".csv"))
}
```
